---
title: "hw4-timothy-welch"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(formattable)
library(plotly)
library(lubridate)
library(RColorBrewer)
```


### Question 1: Monthly Payment of Amortization Formula

#### 1.1) Function amortization()
```{r}
#' Calculate periodic payments of a loan
#' Based on interest rate, number of years, number of periods per year, and principal of loan
#' Periods must either by 0 or positive integer

amortization <- function(principal = 1000, rate = 0.03, years = 30, periods = 12){
  if(periods <=0){
    stop('periods must be positive')
  }
  else{
    pr = rate/periods
    times = years * periods
    principal * ((pr)*((1 + pr)^(times)))/(((1 + pr)^(times))-1)
  }
}
```

**Periodic Payment**

*r = interest rate,*
*P = loan principal,*
*k = payment periods per year,*
*n = number of years*

\(pmt = P * \frac{\frac{r}{k}(1 + \frac{r}{k})^{nk}}{(1+ \frac{r}{k})^{nk} - 1}\)


```{r}
#testing Amortization function

amortization(1000, 0.03, 2, 12)
amortization(1000, 0.03, 2, -6)
```
#### 1.2) Use amortization()

```{r}
#vector of possible years, each index position is # of years
n = c(seq(1:100))
#empty vector
vec = c()
#appending each calculation to empty vector
for(elem in n){
  vec = append(vec, amortization(45000, 0.0525, years = elem, 12))
}
#converting to integer to properly search
vec = as.integer(vec)
#searching to find years
years = which(vec == 200)
years
```

It will take about 75 years of paying $200 per month to fully pay off the principal of the loan:

```{r}
amortization(45000, 0.0525, 10, 12)
```

To pay off the loan in 10 years, it will require paying $482.81 per month. 

### Question 2: Loan Amortization Calculator

#### 2.3) Function loan_calculator()
```{r}

#' Calculates periodic payments of a loan
#' Based on interest rate, number of years, number of periods per year, and principal of loan
#' Periods must either by 0 or positive integer

princ_bal <- function(payment = 200, term, principal = 1000, rate = 0.03, periods = 12){
  if(periods <=0){
    stop('periods must be positive')
  }
  else{
    pr = rate/periods
    principal * ((1 + pr)^term) - ((payment*((1 + pr)^term - 1))/pr)
  }
}


#' Creates loan amortization data-frame, a time-line of loan payments for lifespan of loan
#' Data frame has 5 columns: term of payment, balance, payment amount, interest paid in the term, and principal paid in the term
#' Inputs = principal amount, number of years, interest rate, and number of periods per year
#' Periods must either by 0 or positive integer

loan_calculater <- function(principal = 10000, rate = 0.03, years = 10, periods = 12){
  #term variable
  term = c(seq(0,(years*periods)-1)) 
  
  #payment variable
  pay = amortization(principal, rate, years, periods) 
  payment = c(rep(pay, length(term)))
  
  #balance variable
  balance = c(rep(principal, length(term))) 
  for(pos in 2:length(balance)){
  balance[pos] = princ_bal(pay, term[pos], principal, rate)
  }
  
  #interest variable
  interest = c(rep(0, length(term)))
  for(pos in 1:length(interest)){
  interest[pos] = balance[pos] * rate/periods
  }
  
  #principal variable
  principal = c(rep(0, length(term)))
  for(pos in 1:length(principal)){
  principal[pos] = pay - interest[pos]
  }
  
  #creating data frame
  amort_df = data.frame(term = term, balance = balance, payment = payment, interest = interest, principal= principal)
  amort_df
}

```

##### Variables:

*r = interest rate,*
*P = loan principal,*
*k = payment periods per year,*
*n = number of years,*
*m = payment term*

##### Formulas:

**Periodic Payment**

\(pmt = P * \frac{\frac{r}{k}(1 + \frac{r}{k})^{nk}}{(1+ \frac{r}{k})^{nk} - 1}\)

**Principal Balance**

\(balance_{m} = P * (1 + \frac{r}{k})^m - pmt * \frac{(1 + \frac{r}{k})^{m} - 1}{\frac{r}{k}}\)

**Interest Paid**

\(interest_{m} = balance_{m} * \frac{r}{k}\)

**Principal Paid**

\(principal_{m} = pmt - interest_{m}\)

#### 2.4) Testing loan_calculator()

```{r}
loan_calculater(10000, 0.02, 1, 12)
```


### Question 3: Loan Amortization Chart

#### 3.1) Part 1
```{r}
bros <- loan_calculater(100000, 0.075, 30, 12)

year = c(rep(0, nrow(bros)))
years = c(seq(2011,2039))

#indexing outlier years manually
year[seq(0:9)] = 2010
year[seq(length(year) - 3, length(year))] = 2040

#for loop to fill in rest of years
for(pos in 1:length(years)){
  position = pos*10+(2*(pos-1))
  year[seq(position, position + 11)] = years[pos]
}
#adding year vector to df
bros$Date = year

#converting df to grouped by year
mort_df<- bros %>%
  group_by(Date)%>%
  summarise(Principal = sum(principal), Interest = sum(interest))

#creating balance by year vector
Balance = c(rep(0, nrow(mort_df)))
Balance[1] = bros$balance[10]

for(pos in 2:length(Balance) - 1){
  position = pos*10+(2*(pos-1))
  Balance[pos] = bros$balance[position]
}

#adding to df
mort_df$Balance = currency(Balance)

#displaying df
head(mort_df)
tail(mort_df)
```
#### 3.1) Part 2: ggplotly graphic

```{r}
#preparing data
loan <- loan_calculater(100000, 0.075, 30, 12)

#adding date column
loan$Date = seq(my('04-2010'), my('03-2040'), by = "month")

#converting to currency
loan$Principal <- currency(loan$principal)
loan$Balance <- currency(loan$balance)
loan$Interest <- currency(loan$interest)
```


```{r}
#visualization

labs <- c('Balance (x100)', 'Principal Payment', 'Interest Payment')
fill <- c("#CAB2D6" , "#B2DF8A", "#FB9A99" )
x <- as.Date(c('2020-01-01', '2035-01-01', '2012-01-01'))
y <- c(950, 750, 700)
labeldata <- data.frame(x, y, labs, fill)

amort_vis<- loan %>%
  ggplot(aes(x = Date)) +
  geom_area(aes(y = Principal), fill="#33A02C", alpha=0.2) +
  geom_point(aes(y = Principal), color="#33A02C", size = 0.25) +
  geom_area(aes(y = Interest), fill="#E31A1C" , alpha=0.2) +
  geom_point(aes(y = Interest), color="#E31A1C" , size = 0.25) +
  geom_point(aes(y = Balance/100), color = "#6A3D9A", size = 0.25) +
  theme_classic() +
  theme(text = element_text(family = 'Times New Roman'))+
  labs(title = 'Payment Breakdown by Month of a $100,000, 7.5%, 30 Year Loan',
       y = ' $',
       x = 'Date')+
  geom_label(aes(x, y, label = labs, group = NULL), fill = fill, data = labeldata, size = 4, family = 'Times New Roman') 

#static visualization in ggplot
amort_vis

#interactive plotly visualization
plot <- ggplotly(amort_vis)
plot
```

